---
title: "Ontario's At-Risk Bumble Bees"
subtitle: "An open-data screen of range, decline, and survey gaps — Quadrat Studio"
author: "Susan Frye · Quadrat Studio"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    number-sections: true
    fig-width: 8
    fig-height: 4.5
    code-fold: true
    embed-resources: true
    include-in-header:
      text: |
        <link href="https://fonts.googleapis.com/css2?family=Space+Grotesk:wght@500;700&family=Inter:wght@400;500&display=swap" rel="stylesheet">
        <style>
          body { font-family: 'Inter', sans-serif; }
          h1, h2, h3, h4, .title, .subtitle { font-family: 'Space Grotesk', sans-serif; letter-spacing: -0.02em; }
          a { color: #0d9488; }
          #title-block-header .quarto-title-banner { background: #0a0a0a; }
        </style>
params:
  region: "Ontario"
  year_start: 2000
  year_end: 2024
  recent_from: 2015
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
# SETUP (run once): install.packages(c("rgbif","dplyr","ggplot2","leaflet","knitr","DT","scales","tibble"))
library(rgbif); library(dplyr); library(ggplot2); library(leaflet)
library(knitr); library(DT); library(scales); library(tibble)

teal <- "#0d9488"; okabe <- c("#D55E00", "#E69F00", "#009E73", "#56B4E9")

# ── Focal species and status (same config as the dashboard app) ──────────────
bee_meta <- tibble::tribble(
  ~sci,                    ~common,                     ~status,            ~role,
  "Bombus affinis",        "Rusty-patched Bumble Bee",  "Endangered",       "focal",
  "Bombus terricola",      "Yellow-banded Bumble Bee",  "Special Concern",  "hero",
  "Bombus pensylvanicus",  "American Bumble Bee",       "Special Concern",  "focal",
  "Bombus fervidus",       "Yellow Bumble Bee",         "Special Concern",  "focal",
  "Bombus impatiens",      "Common Eastern Bumble Bee", "Secure",           "baseline"
)
at_risk_status <- c("Endangered", "Special Concern")
hero_common     <- bee_meta$common[bee_meta$role == "hero"]
baseline_common <- bee_meta$common[bee_meta$role == "baseline"]

# ── Build & cache (self-contained, reproducible) ─────────────────────────────
# Three cheap pulls: exact yearly Bombus totals (the proportion denominator) and
# exact yearly per-species counts via GBIF facets, plus focal-species coordinates.
# No naive count-trend is used anywhere; see the methods notes inline.
cache_file <- "cache_report_bombus.rds"

facet_year <- function(sci = NULL, key = NULL) {
  args <- list(country = "CA", stateProvince = params$region,
               year = paste0(params$year_start, ",", params$year_end),
               hasCoordinate = TRUE, limit = 0, facet = "year", facetLimit = 200)
  if (!is.null(sci)) args$scientificName <- sci
  if (!is.null(key)) args$taxonKey <- key
  f <- do.call(occ_search, args)$facets$year
  if (is.null(f) || nrow(f) == 0) return(tibble(year = integer(), n = integer()))
  tibble(year = as.integer(f$name), n = as.integer(f$count))
}

if (!file.exists(cache_file)) {
  key <- name_backbone("Bombus")$usageKey
  denom_year <- facet_year(key = key) |> rename(total = n)
  sp_year <- bind_rows(lapply(seq_len(nrow(bee_meta)), function(i) {
    r <- bee_meta[i, ]; fy <- facet_year(sci = r$sci); if (nrow(fy) == 0) return(NULL)
    fy$common <- r$common; fy$status <- r$status; fy
  }))
  focal <- bind_rows(lapply(seq_len(nrow(bee_meta)), function(i) {
    r <- bee_meta[i, ]
    res <- tryCatch(occ_search(scientificName = r$sci, country = "CA",
      stateProvince = params$region,
      year = paste0(params$year_start, ",", params$year_end), hasCoordinate = TRUE,
      limit = 5000,
      fields = c("decimalLatitude","decimalLongitude","year","month"))$data,
      error = function(e) NULL)
    if (is.null(res) || nrow(res) == 0) return(NULL)
    res$common <- r$common; res$status <- r$status; res
  })) |>
    filter(!is.na(decimalLatitude), !is.na(decimalLongitude), !is.na(year)) |>
    rename(lat = decimalLatitude, lon = decimalLongitude) |>
    mutate(decade = paste0(floor(year/10)*10, "s"), at_risk = status %in% at_risk_status)
  saveRDS(list(denom_year = denom_year, sp_year = sp_year, focal = focal), cache_file)
}
d <- readRDS(cache_file)
denom_year <- d$denom_year; sp_year <- d$sp_year; focal <- d$focal

n_total <- sum(denom_year$total)
hero <- focal |> filter(common == hero_common)
base_lat <- mean(focal$lat[focal$common == baseline_common])
hero_lat <- mean(hero$lat)
north_km <- round((hero_lat - base_lat) * 111)
```

## Why this report exists {#sec-why}

[Wildlife Preservation Canada](https://wildlifepreservation.ca/bumble-bee-recovery/)'s
Bumble Bee Recovery program has reported two patterns from its Ontario field work: the
**Yellow-banded Bumble Bee** (*Bombus terricola*) turning up further **north**, and the
Endangered **Rusty-patched Bumble Bee** (*Bombus affinis*) — once widespread in southern
Ontario — proving impossible to relocate despite intensive annual searches.

This report asks how much of that is visible in **open data**, using
`r comma(n_total)` *Bombus* records for `r params$region`
(`r params$year_start`–`r params$year_end`) from [GBIF.org](https://www.gbif.org),
which aggregates iNaturalist, Bumble Bee Watch, museum, and research records.

::: callout-important
**Occurrence records measure observer effort, not just bees.** Records of almost every
species rise over time because more people record — and that boom is concentrated in the
populous south. So this report deliberately **avoids raw count-trends**, including raw
range-shift trends, and leans on metrics that survive the bias: a same-period baseline
comparison, the proportion of records, and a survey-effort grid. Absence of records is
**not** evidence of absence.
:::

## A northern bee — but read it carefully {#sec-north}

It is tempting to track a species' mean latitude over time and call a rise a "northward
shift." In this data that fails: the Yellow-banded centroid actually drifts **south**,
dragged by the recent explosion of southern-Ontario recorders — a textbook effort
artefact, not biology.

```{r centroid-trap}
hero |> group_by(decade) |>
  summarise(`Mean latitude (°N)` = round(mean(lat), 2), Records = n(), .groups = "drop") |>
  kable(caption = "Raw mean latitude of the Yellow-banded Bumble Bee by decade — it drifts *south*, because recent recording is concentrated in the populous south. This is why a raw range-shift trend is not trustworthy here.")
```

What *is* defensible is a **same-period, same-observer** comparison: where does each
at-risk species sit relative to the common bee that everyone records? Bees north of that
baseline are genuinely more northern in range, regardless of where people happen to look.

```{r latitude-vs-baseline}
lat_tab <- focal |> group_by(common, status) |>
  summarise(mean_lat = mean(lat), sd = sd(lat), n = n(), .groups = "drop") |>
  arrange(mean_lat)
ggplot(lat_tab, aes(mean_lat, reorder(common, mean_lat), colour = status)) +
  geom_vline(xintercept = base_lat, linetype = "dashed", colour = "#0a0a0a") +
  geom_point(size = 3.5) +
  annotate("text", x = base_lat, y = 0.6, label = "common-bee baseline",
           hjust = -0.05, size = 3.2) +
  scale_colour_manual(values = c("Endangered" = okabe[1], "Special Concern" = okabe[2],
                                 "Secure" = okabe[3])) +
  labs(x = "Mean latitude (°N)", y = NULL, colour = NULL,
       title = "Where each species is recorded, vs the common baseline bee") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

The Yellow-banded Bumble Bee is recorded on average **`r abs(north_km)` km
`r ifelse(north_km >= 0, "north", "south")`** of the common Eastern Bumble Bee in the
same period — a northern-range signal that holds up against the southern recording boom.
A genuine *temporal* shift, though, can only be established with structured field
data — exactly the work WPC does.

## The decline signal {#sec-decline}

Expressing each species as a **share of all *Bombus* records** in the same year divides
out the year-to-year change in total effort. A falling share is a decline signal that a
rising raw count would hide.

```{r proportion}
prop <- sp_year |>
  left_join(denom_year, by = "year") |>
  mutate(pct = 100 * n / total) |>
  filter(total >= 20)
ggplot(prop, aes(year, pct, colour = common)) +
  geom_line(linewidth = 0.9) + geom_point(size = 1.3) +
  scale_colour_brewer(palette = "Dark2") +
  labs(x = NULL, y = "Share of all Bombus records (%)", colour = NULL,
       title = "At-risk species as a share of Ontario Bombus records") +
  theme_minimal(base_size = 13) + theme(legend.position = "bottom")
```

```{r affinis-note}
affinis_total  <- sum(sp_year$n[sp_year$common == "Rusty-patched Bumble Bee"])
affinis_recent <- focal |>
  filter(common == "Rusty-patched Bumble Bee", year >= params$recent_from) |> nrow()
```

The **Rusty-patched Bumble Bee** accounts for **`r affinis_total` record(s)** in
`r params$region` across the whole period, and **`r affinis_recent`** with coordinates
since `r params$recent_from` — the open-data echo of WPC's inability to relocate the
species in the field.

## Where to survey next {#sec-gaps}

Rather than claim absence, we bin records into 0.5° cells and ask, for each cell, whether
an at-risk bee has been recorded *recently*, only *historically*, or *never* despite
other bumble bees being present. The amber and blue cells are the highest-value targets
for directed survey effort.

```{r gap-map}
g <- focal |>
  mutate(gx = floor(lon/0.5)*0.5, gy = floor(lat/0.5)*0.5) |>
  group_by(gx, gy) |>
  summarise(total = n(), risk_any = sum(at_risk),
            risk_recent = sum(at_risk & year >= params$recent_from), .groups = "drop") |>
  mutate(class = case_when(
    risk_recent > 0 ~ "At-risk present (recent)",
    risk_any   > 0  ~ "At-risk historic only",
    TRUE            ~ "Bombus recorded, no at-risk"))
gpal <- colorFactor(c(okabe[3], okabe[2], okabe[4]),
  levels = c("At-risk present (recent)","At-risk historic only","Bombus recorded, no at-risk"))
leaflet(g) |> addProviderTiles(providers$CartoDB.Positron) |>
  addRectangles(~gx, ~gy, ~gx+0.5, ~gy+0.5, weight = 0.5, color = "#fff",
    fillColor = ~gpal(class), fillOpacity = ~pmin(0.85, 0.25 + total/max(g$total)),
    popup = ~paste0(class, "<br>", total, " records; ", risk_any, " at-risk")) |>
  addLegend("bottomright", pal = gpal,
    values = c("At-risk present (recent)","At-risk historic only","Bombus recorded, no at-risk"),
    title = "0.5° cells", opacity = 0.9)
```

```{r gap-counts}
hist_only <- sum(g$class == "At-risk historic only")
none_yet  <- sum(g$class == "Bombus recorded, no at-risk")
```

There are **`r hist_only` cells** with at-risk bumble bees on record but none since
`r params$recent_from`, and a further **`r none_yet` well-surveyed cells** that have never
recorded an at-risk species — a concrete, rankable shortlist for field planning.

## Interpretation & caveats {#sec-caveats}

- **The signals are suggestive, not proof.** A same-period baseline comparison controls
  for *where* people record, but not for every confound; photo-based IDs carry error.
- **Northern Ontario is under-sampled.** Sparse records there are as likely to reflect no
  observers as no bees.
- **No clean temporal range shift can be drawn from this opportunistic data** — that
  requires structured monitoring. This screen is built to *prioritise* such effort.
- **A recovery-grade analysis** would model detection effort explicitly and fold in
  targeted field surveys.

---

*Prepared with Quadrat Studio · Susan Frye · [quadratstudio.ca](https://quadratstudio.ca).
Reproducible source: this `.qmd` re-renders from public GBIF data.*
