---
title: "Biodiversity Monitoring Summary"
subtitle: "A reproducible reporting template — Quadrat Studio"
author: "Susan Frye · Quadrat Studio"
date: today
format:
  html:
    theme: cosmo
    toc: true
    toc-location: left
    number-sections: true
    fig-width: 8
    fig-height: 4.5
    code-fold: true
    embed-resources: true
    include-in-header:
      text: |
        <link href="https://fonts.googleapis.com/css2?family=Space+Grotesk:wght@500;700&family=Inter:wght@400;500&display=swap" rel="stylesheet">
        <style>
          body { font-family: 'Inter', sans-serif; }
          h1, h2, h3, h4, .title, .subtitle { font-family: 'Space Grotesk', sans-serif; letter-spacing: -0.02em; }
          a { color: #0d9488; }
          #title-block-header .quarto-title-banner { background: #0a0a0a; }
        </style>
params:
  region: "Ontario"
  year_start: 2000
  year_end: 2024
  taxa:
    - "Danaus plexippus"
    - "Bombus affinis"
    - "Bombus terricola"
    - "Dolichonyx oryzivorus"
    - "Emydoidea blandingii"
execute:
  warning: false
  message: false
---

```{r setup}
#| include: false
# SETUP (run once): install.packages(c("rgbif","dplyr","ggplot2","leaflet","knitr","DT","scales"))
library(rgbif)
library(dplyr)
library(ggplot2)
library(leaflet)
library(knitr)
library(DT)
library(scales)

# ── Download & cache ────────────────────────────────────────────────────────
# This report is fully reproducible: rendering it re-pulls (or re-uses a cache
# of) public GBIF records, so the same code produces the same document months
# later. Swap `params$taxa` / region for a different study and re-render.
cache_file <- "cache_report_gbif.rds"

if (!file.exists(cache_file)) {
  raw <- lapply(params$taxa, function(sci) {
    tryCatch({
      res <- occ_search(
        scientificName = sci,
        country        = "CA",
        stateProvince  = params$region,
        year           = paste0(params$year_start, ",", params$year_end),
        limit          = 1000,
        hasCoordinate  = TRUE,
        fields         = c("decimalLatitude", "decimalLongitude",
                           "year", "month", "species")
      )$data
      if (!is.null(res) && nrow(res) > 0) res$query_sci <- sci
      res
    }, error = function(e) NULL)
  })
  occ <- bind_rows(Filter(Negate(is.null), raw)) |>
    filter(!is.na(decimalLatitude), !is.na(decimalLongitude), !is.na(year)) |>
    rename(lat = decimalLatitude, lon = decimalLongitude) |>
    mutate(species = ifelse(is.na(species), query_sci, species),
           decade  = paste0(floor(year / 10) * 10, "s"))
  saveRDS(occ, cache_file)
}
occ <- readRDS(cache_file)

teal <- "#0d9488"
```

## Scope and purpose {#sec-scope}

This template demonstrates the kind of **defensible, reproducible reporting**
Quadrat Studio delivers to conservation authorities, municipalities, and
environmental consultants. It summarises publicly available occurrence records
for a set of focal species in **`r params$region`** over
**`r params$year_start`–`r params$year_end`**, drawn from
[GBIF.org](https://www.gbif.org).

Everything below is generated from code: the figures, tables, and numbers update
automatically when the underlying data or parameters change. For a real
engagement, the same machinery runs on *your* monitoring data and produces a
branded report ready for a regulator or board.

::: callout-note
**This is an illustrative template using open data.** Occurrence records reflect
sampling effort as much as true distribution; absence of records is not evidence
of absence. A project report would document data provenance, QA/QC, and analysis
assumptions in full.
:::

## Data summary {#sec-summary}

```{r overview}
n_records <- nrow(occ)
n_species <- length(unique(occ$species))
yr_span   <- paste0(min(occ$year), "–", max(occ$year))
```

The compiled dataset contains **`r comma(n_records)` records** across
**`r n_species` species**, spanning **`r yr_span`**.

```{r species-table}
occ |>
  group_by(Species = species) |>
  summarise(Records = n(),
            `First record` = min(year),
            `Most recent`  = max(year),
            .groups = "drop") |>
  arrange(desc(Records)) |>
  datatable(rownames = FALSE, options = list(dom = "tp", pageLength = 10),
            caption = "Records per focal species.")
```

## Spatial distribution {#sec-map}

```{r map}
pal <- colorFactor("viridis", domain = occ$species)
leaflet(occ) |>
  addProviderTiles(providers$CartoDB.Positron) |>
  addCircleMarkers(~lon, ~lat, radius = 3, stroke = FALSE, fillOpacity = 0.6,
                   color = ~pal(species),
                   popup = ~paste0("<b>", species, "</b><br>", year)) |>
  addLegend("bottomright", pal = pal, values = ~species, title = "Species",
            opacity = 0.9)
```

## Records through time {#sec-trend}

```{r trend}
occ |>
  count(decade) |>
  ggplot(aes(decade, n)) +
  geom_col(fill = teal) +
  labs(x = NULL, y = "Records", title = "Records by decade") +
  theme_minimal(base_size = 13)
```

## Seasonal detection (phenology) {#sec-phenology}

```{r phenology}
occ |>
  filter(!is.na(month)) |>
  count(month) |>
  ggplot(aes(factor(month, levels = 1:12, labels = month.abb), n, group = 1)) +
  geom_col(fill = teal) +
  labs(x = NULL, y = "Records", title = "Records by month — when species are detected") +
  theme_minimal(base_size = 13)
```

The phenology profile indicates **when** records accumulate through the year —
directly useful for scheduling field surveys so effort lands in the detection
window for target species.

## Interpretation & next steps {#sec-next}

- **Survey timing.** Concentrate monitoring effort in the months with peak
  detections (see @sec-phenology).
- **Spatial gaps.** Areas with few records (@sec-map) may reflect under-sampling
  rather than true absence — candidates for targeted survey.
- **Trend caution.** Apparent changes over time (@sec-trend) are confounded with
  changes in observer effort; a defensible trend analysis would model effort
  explicitly.

---

*Prepared with Quadrat Studio · Susan Frye · [quadratstudio.ca](https://quadratstudio.ca).
Reproducible source: this `.qmd` re-renders from public GBIF data.*
